Left-linear rewriting in adhesive categories

### 2024-04-24 concur/fusione.tex
typos
minori

\begin{abstract}
  When can two sequential steps performed by a computing device be
  considered (causally) independent?  This is a relevant question for
  concurrent and distributed systems, since independence means that
  they could be executed in any order, and potentially in
  parallel. Equivalences identifying rewriting sequences which differ
  only for independent steps are at the core of the theory of
  concurrency of many formalisms.
  % 
  We investigate the issue in the context of the double pushout
  approach to rewriting in the general setting of adhesive categories.
  %, a very general setting for rewriting.  
  While a consolidated theory exists for linear rules, which
  can consume, preserve and generate entities, this paper focuses on
  left-linear rules which may also ``merge'' parts of the state.
  % 
  An apparently minimal, yet technically hard enhancement, 
  % 
  % We refine the established notion of sequential independence
  % between steps, and ...
  since % the canonical notion of independence used for linear rewriting
  % systems does not work out of the box, but it requires the
  % restriction to well-behaved classes of left-linear rules.
  % %
  % Moreover,
  a standard characterisation of independence that -- in the
  linear case -- allows one to derive a number of properties,
  essential in the development of a theory of concurrency, no longer
  holds.
  %and this requires a substantial revision of the theory.
  %
  The paper performs an in-depth study of the notion of independence
  for left-linear rules: it introduces a novel characterisation of
  independence, identifies well-behaved classes of left-linear
  rewriting systems, and provides some fundamental results including a
  Church-Rosser property and the existence of canonical equivalence
  proofs for concurrent computations. These results properly extends
  the class of formalisms that can be modelled in the adhesive
  framework.
\end{abstract}

@@

Among rewriting-based formalisms, those manipulating graph-like
structures proved useful in many settings for representing the
dynamics of distributed systems: the graph is used to represent the
entities and their relations within states, while rewriting steps,
modifying the graph, model computation steps that result in changes to
the system state. The DPO (double-pushout) approach~\cite{EhrigPS73}
is nowadays considered a standard one for these structures, thanks to
its locality (rule application has a local effect on a state,
differently from more recent approaches like
SqPO~\cite{CorradiniHHK06} or PBPO$^+$~\cite{OverbeekER23}) and to its
flexibility: the rules allow to specify that, in a rewriting step,
some entities in the current state are removed, some others are needed
for the rewriting step to occur but remain unaltered, and some new
entities can be generated. Concerning concurrency, the chosen notion
is switch (also referred to as shift) equivalence, and independence is formally captured by what is
called the Church-Rosser theorem of DPO
rewriting~\cite[Chapter 3, Section~3.4]{CorradiniMREHL97}. %cite[Section~4]{Ehrig78}

@@

As an initial step, % in performing this exploration, 
we observe that for
left-linear rules some fundamental properties that play a role
in the concurrency theory of linear rewriting need to be deeply revised, or
plainly do not hold.
%
Firstly, the notion of sequential independence and the Church-Rosser
theorem, ensuring that sequential independent steps can be performed
in reverse order, does not hold out of the box. Even worse, given two
independent steps it might be possible to switch them in several
different ways.
%
As we will observe, this prevents the development of a sensible theory
of concurrency, since switches are used as a basis for equating
sequences of rewriting steps that differ only for the order of
independent steps.

@@

On the one hand, they capture most categories of
interest for rewriting. Indeed, we identify quite general classes of
left-linear rewriting systems on graph-like structures that are
well-switching.
% and we show that rewriting systems on graphs with
% equivalences are so.
As a sanity check, we prove that linear rewriting systems on
$\mathcal{M}$-adhesive categories are well-switching.

On the other hand, we argue that a concurrency theory of rewriting can
be developed for well-switching rewriting systems, recovering, sometimes in
weakened form, results that hold in the linear case.
% Specifically, the Church-Rosser theorem is fully recovered.
%
The development has to face a main technical obstacle. In the
literature on linear rewriting systems, a central result is the
characterisation of switch equivalence of sequences of steps based on
what are called processes or consistent
permutations~\cite{CMR:GP,Handbook,BaldanCHKS06,heindel2009category}.
In turn this is the key to derive some
% basic but
fundamental
properties for linear rewriting,
% systems
namely

@@

    A $\mor(\X)$-Van Kampen square will be called simply \emph{Van
      Kampen} and similarly a $\mor(\X)$-stable square will be called
    \emph{stable}.
  \end{definition}

@@

$\mathcal{M}$-adhesivity is well-behaved with respect to  the construction of slice and functor categories \cite{mac2013categories}, as shown by the following theorems~\cite{ehrig2006fundamentals,lack2005adhesive}.

@@

\begin{example}
  \label{ex:seq-ind}
  Consider a rewriting system in $\cat{Graph}$, the category of
  directed graphs and graph morphisms, which is adhesive. The set of
  rules $\R = \{ \rho_0, \rho_1, \rho_2\}$ is in
  Fig.~\ref{fi:rules}, where numbers are used to represent the
  morphisms from the interface to the left- and right-hand
  sides. Rules $\rho_0$ and $\rho_1$ are linear: $\rho_0$ generates a
  new node and a new edge, while $\rho_1$ creates a self-loops edge. Instead,
  $\rho_2$ is left-linear but not linear, as it ``merges'' two nodes.

@@

We can identify a large class of adhesive categories such that all
left-linear rewriting systems over such categories are strong
enforcing. Such class includes $\cat{Set}$ and it is closed
by comma and functor category constructions.  As such, it includes
essentially all categories (e.g., presheaves over set) that are
typically considered for modelling purposes. Notably, it contains the
category $\cat{Graph}$ of directed graphs.
This is a natural generalisation from adhesive to $\mathcal{M}$-adhesive categories of a class studied in \cite{baldan2011adhesivity} (see \Cref{app:fill} for details).

@@

\begin{example}[Non-strong enforcing left-linear rewriting system]
  \label{ex:diff2}
  In light of \Cref{pr:church}, \Cref{ex:diff1} provides an example of
  an independence pair which is not strong. This gives an example of a
  left-linear rewriting system in a $\mathcal{M}$-adhesive category
  pathological since $\mathcal{M}=\mathsf{Iso(\X)}$, but this is expected, 
  since all natural examples seem to belong to the well-behaved class mentioned above. 
 % .\todo{P: Scritto qualcosa, probabilmente da migliorare \\ D: per me va bene così}
%  pathological since $\mathcal{M}=\mathsf{Iso(\X)}$, but it needs to be so, since natural examples are all in the well-behaved class mentioned above. 
 % .\todo{P: Scritto qualcosa, probabilmente da migliorare \\ D: per me va bene così}
  % Instead every
  % left-linear system in category $\cat{Graph}$ is strong enforcing, by
  % the results of Appendix~\ref{app:fill}, summarized
  % in~\Cref{ex:graph-type-bb} and~\Cref{cor:llr-is-strEnf}.
\end{example}

@@

  As a consequence, $\der{D}$ and $\der{D}'$ would be switch
  equivalent, but this is not acceptable when viewing equivalence
  classes of derivations as concurrent computations: in $\der{D}$ the
  first two steps are not sequential independent, while in $\der{D}'$
  they are, intuitively because in $\der{D}$ rule $\rho_1$ uses the
  node generated by $\rho_0$ (adding a self loop to it), while in
  $\der{D}'$ rule $\rho_1$ uses the node that was in the initial
  graph. Also observe that the graphs $G_2$ and $G_2'$ produced after
  two steps in $\der{D}$ and $\der{D'}$ are not isomorphic. From the
  technical point of view, the property of being switch equivalent is
  not closed by prefix, and this prevents to derive a sensible
  concurrent semantics: In fact
  $\der{D} = \dder{D}_0\cdot \dder{D}_1 \cdot \dder{D}_2$ and
  $\der{D}' = \dder{D}_0'\cdot \dder{D}_1' \cdot \dder{D}_2'$ are
  switch equivalent, while if we consider the first two steps,
  derivations $\dder{D}_0 \cdot \dder{D}_1$ and
  $\dder{D}_0' \cdot \dder{D}_1'$ are not switch equivalent.
  % if we limit to the first
  %two steps, we obtain derivations $\dder{D}_0 \cdot \dder{D}_1$ and
  %$\dder{D}_0' \cdot \dder{D}_1'$ that are not switch equivalent, while
  %$\der{D} = \dder{D}_0\cdot \dder{D}_1 \cdot \dder{D}_2$ and
  %$\der{D}' = \dder{D}_0'\cdot \dder{D}_1' \cdot \dder{D}_2'$ would
  %be.
\end{example}

@@

A first class consists of those rewriting systems over
possibly hierarchical graphical structures obtained as algebras of
suitable signatures where rules are constrained not to merge elements
of top level sorts in the hierarchy (for graphs, nodes can be
merged while edges cannot). The idea here is to consider graph
structures as presheaves on categories in which there are objects that play 
the role of \emph{roots}, i.e.~objects that are not the codomain of 
any arrow besides the identity. 
%\todo{D: quest'ultima frase ha senso?}
%We first introduce signatures that are
%essentially acyclic graphs whose nodes can be seen as sorts.

\begin{definition}[Root-preserving graphical rewriting systems]
	Let $\X$ be a category, an object $X\in \X$ is a \emph{root} if the only arrow with codomain $X$ is $\id{X}$.
	The category $\gph{X}$ of \emph{$\X$-graphs} is the category
	$\Set^{\X}$. A
	\emph{root-preserving graphical rewriting system} is a left-linear
	rewriting system $(\gph{X}, \R)$ such that for each rule
	$(l\colon K\to L, r\colon K\to R)$ in $\R$ it holds
	\begin{enumerate}
		\item for every $X\in \X$ and $x\in L(X)$, there exists a root $Y$
		 and an arrow $f\colon Y\to X$ such that $x$ belongs to the image of
		$L(f)\colon L(Y)\to L(X)$;
		\item $r\colon K\to R$ is mono on the roots, i.e.~for every root $X\in \X$ the component $r_X:K(X)\to R(X)$ is injective.
	\end{enumerate}
\end{definition}

@@

For instance, the category $\cat{Graph}$ can be obtained
by taking as $\X$ the category generated by $E \rightrightarrows V$. In this case $E$ is the only root, hence, condition $1$ asks that in the left-hand side of each rule
there are no isolated nodes, while condition $2$ asks that the
morphism $r: K \to R$ is injective on edges, i.e.~it can only merge nodes.

@@

Left-linear (w.r.t.~$\reg(\cat{EGraphs})$) rewriting systems on $\cat{EGraphs}$ where the right half of a rule might be mono but not regular 
are well-switching. They have been used in~\cite{BaldanGM06} for the graphical implementation of nominal calculi, where,
differently from~\cite{Gad07}, as a result of name passing the received name is not merged with a local one at the receiver,
but put instead in the same equivalence class, thus better tracing the causal dependencies among processes reductions.
\end{example}

@@

\begin{example}
  \label{ex:non-global}
  Consider a new rewriting system in $\cat{Graph}$ consisting of the
  rules $\sigma_0$, $\sigma_1$, and $\sigma_2$ in
  Fig.~\ref{fi:rules-cas}.
  %\todo{P: rev.2 dice ``In the top-left and
   % top-middle nodes of $\rho_3$, the labels "12" should be "1" and
   % "2".''. Non e' corretto, ma magari si capisce poco ...}  
   Consider the derivation $\der{F}$ in Fig.~\ref{fi:derF}. Note that rule
  $\sigma_2$ needs the black self-loop on node $12$ to be applied. This can be
  generated either by $\sigma_0$ or by $\sigma_1$ merging nodes $1$ and
  $2$, hence at least one of them must precede the application of
  $\sigma_2$. Indeed, in $\der{F}$ it can be easily seen that the second
  and the third step applying $\sigma_1$ and $\sigma_2$ are
  independent. However, we could switch twice the application of
  $\sigma_0$, bringing it in the last position, obtaining a derivation
  $\der{F}'$ as depicted in Fig.~\ref{fi:derF1}, where the
  applications of $\sigma_1$ and $\sigma_2$ are no longer independent.

@@

The lemmas above open the way to the key result of this section, which establishes a canonical form for
switching derivations.

The first result shows that when equating two derivation sequences by performing a series of switchings, we can limit ourselves to switchings that inverte the order of steps that in the target derivation has to be reversed, i.e.~we can limit ourselves to applying inversions.

@@

While inversions can be
performed in any order in the case of linear rewriting systems, this is no longer true for left-linear
rewriting systems due to the fact that, as already observed, globality of independence holds in a weak form. 
For instance, if we get back to Example~\ref{ex:non-global} and consider derivation $\der{F}$ in Fig.~\ref{fi:derF}, 
the last two steps applying rules $\sigma_1$ and $\sigma_2$ can be switched, thus leading to a derivation, 
call it $\der{F}''$,  applying $\sigma_0, \sigma_2, \sigma_1$. Such $\der{F}''$ is switch equivalent to $\der{F}'$ in Fig.~\ref{fi:derF1}, 
hence starting from $\der{F}''$ we can obtain $\der{F}'$ via a switching sequence, but we cannot start 
by switching $\sigma_0$ and $\sigma_2$, despite the fact that this is an inversion.

@@

The class of well-switching $\mathcal{M}$-adhesive rewriting systems is large. Generalising a result in~\cite{baldan2011adhesivity},
we showed that one of its defining properties (the fact that sequential independence implies switchability) holds 
for a general class of $\mathcal{M}$-adhesive categories that include all presheaves over $\cat{Set}$. Moreover, the second property, uniqueness 
of switching, is ensured when rewriting possibly hierarchical graphical structures, as long as elements belonging to the sorts 
of the roots  are not merged (e.g., one can consider rewriting systems over directed graphs where only nodes are merged, as it 
happens in most of the approaches to the modelling of calculi and biological systems that we cited in the introduction, 
as well as in string diagrams).
%\todo{Giusto, refernences?}

@@

On the practical side, the present paper just surveyed the possibility
of modelling e-graphs. This appears to be quite interesting as
concurrent rewriting on such structures is an active area of
research~\cite{abs-2208-06295}. 
%\todo{F: altre citazioni?} 
We plan to make the correspondence precise
and investigate how the analysis techniques enabled by a concurrent
semantics of rewriting can impact on e-graphs.

@@

\begin{definition}[Consistent permutation]
	\label{def:permcon}
	Let $\X$ be an $\mathcal{M}$-adhesive category and $(\X, \R)$ 
	a left-linear rewriting system.  Take two
	non-empty decorated derivations $(\der{D}, \alpha, \omega)$ and
	$(\der{D}', \alpha', \omega')$ with the same length and with
	isomorphic sources and targets.

	If $\der{D}=\{\dder{D}_i\}_{i=0}^n$ and
	$\der{D}'=\{\dder{D}'_i\}_{i=0}^n$ with associated sequence of rules
	$r(\der{D})=\{\rho_i\}_{i=0}^n$ and
	$r(\der{D}')=\{\rho'_i\}_{i=0}^n$, a \emph{consistent permutation}
	between $(\der{D}, \alpha, \omega)$ and
	$(\der{D}', \alpha', \omega')$ is a permutation
	$\sigma\colon [0,n]\to [0,n]$ such that, for every $i\in [0,n]$,
	$\rho_i=\rho'_{\sigma(i)}$ and, moreover, there exists a
	\emph{mediating isomorphism}
	$\xi_\sigma\colon \tpro{D} \to \lpro \der{D}' \rpro$ fitting in the
	following diagrams, where $m_i, m'_i, h_i$ and $h'_i$ are,
	respectively, the matches and co-matches of $\dder{D}_i$ and
	$\dder{D}'_i$

	\[
	\xymatrix@C=30pt{\pi(G_0)\ar[r]^{\alpha} \ar[d]_{\alpha'} & G_0
		\ar[r]^{\iota_{G_0}} &\tpro{D} \ar[d]^{\xi_\sigma}& \pi(G_{n+1})
		\ar[r]^{\omega} \ar[d]_{\omega'} & G_{n+1}
		\ar@{>->}[r]^{\iota_{G_{n+1}}} &\tpro{D} \ar[d]^{\xi_\sigma}\\
		G'_0 \ar[rr]_{\iota'_{G'_0}} & &\lpro \der{D}' \rpro& G'_{n+1}
		\ar@{>->}[rr]_{\iota'_{G'_{n+1}}} & &\lpro \der{D}' \rpro\\L_i
		\ar[r]^{m_i} \ar[d]_{m'_{\sigma(i)}}& G_i \ar[r]^{\iota_{G_i}}
		&\tpro{D} \ar[d]^{\xi_\sigma} & R_i \ar[r]^{h_i}
		\ar[d]_{h'_{\sigma(i)}}& G_{i+1} \ar[r]^{\iota_{G_{i+1}}}
		&\tpro{D} \ar[d]^{\xi_\sigma} \\G'_{\sigma(i)}
		\ar[rr]_{\iota'_{G'_{\sigma(i)}}}&& \lpro \der{D}' \rpro&
		G'_{\sigma(i)+1} \ar[rr]_{\iota'_{G'_{\sigma(i)+1}}}&& \lpro
		\der{D}' \rpro}
	\]
\end{definition}
We can now turn to the proofs of the results of \Cref{subsec:canonical}.

>>>>

\begin{abstract}
  When can two sequential steps performed by a computing device be
  considered (causally) independent?  This is a relevant question for
  concurrent and distributed systems, since independence means that
  they could be executed in any order, and potentially in
  parallel. Equivalences identifying rewriting sequences which differ
  only for independent steps are at the core of the theory of
  concurrency of many formalisms.
  % 
  We investigate the issue in the context of the double pushout
  approach to rewriting in the general setting of adhesive categories.
  %, a very general setting for rewriting.  
  While a consolidated theory exists for so-called linear rules which
  can consume, preserve and generate entities, this paper focuses on
  left-linear rules which may also ``merge'' parts of the state.
  % 
  An apparently minimal, yet technically hard enhancement, 
  % 
  % We refine the established notion of sequential independence
  % between steps, and ...
  since % the canonical notion of independence used for linear rewriting
  % systems does not work out of the box, but it requires the
  % restriction to well-behaved classes of left-linear rules.
  % %
  % Moreover,
  a standard characterisation of independence that -- in the
  linear case -- allows one to derive a number of properties,
  essential in the development of a theory of concurrency, no longer
  holds.
  %and this requires a substantial revision of the theory.
  %
  The paper performs an in-depth study of the notion of independence
  for left-linear rules: it introduces a novel characterisation of
  independence, identifies well-behaved classes of left-linear
  rewriting systems, and provides some fundamental results including a
  Church-Rosser property and the existence of canonical equivalence
  proofs for concurrent computations. These results properly extends
  the class of formalisms that can be modelled in the adhesive
  framework.
\end{abstract}

@@

Among rewriting-based formalisms, those manipulating graph-like
structures proved useful in many settings for representing the
dynamics of distributed systems: the graph serves to represent the
entities and their relations within states, while rewriting steps,
modifying the graph, model computation steps that result in changes to
the system state. The DPO (double-pushout) approach~\cite{EhrigPS73}
is nowadays considered a standard one for these structures, thanks to
its locality (rule application has a local effect on a state,
differently from more recent approaches like
SqPO~\cite{CorradiniHHK06} or PBPO$^+$~\cite{OverbeekER23}) and 10to its
flexibility: the rules allow to specify that, in a rewriting step,
some entities in the current state are removed, some others are needed
for the rewriting step to occur but remain unaltered, and some new
entities can be generated. Concerning concurrency, the chosen notion
is switch (also referred to as shift) equivalence, and independence is formally captured by what is
called the Church-Rosser theorem of DPO
rewriting~\cite[Chapter 3, Section~3.4]{CorradiniMREHL97}. %cite[Section~4]{Ehrig78}

@@

As an initial step, % in performing this exploration, 
we observe that for
left-linear rules a number of fundamental properties that play a role
in the concurrency theory of linear rewriting need to be deeply revised, or
plainly do not hold.
%
Firstly, the notion of sequential independence and the Church-Rosser
theorem, ensuring that sequential independent steps can be performed
in reverse order, does not hold out of the box. Even worse, given two
independent steps it might be possible to switch them in several
different ways.
%
As we will observe, this prevents the development of a sensible theory
of concurrency, since switches are used as a basis for equating
sequences of rewriting steps that differ only for the order of
independent steps.

@@

On the one hand, they capture most categories of
interest for rewriting. Indeed, we identify quite general classes of
left-linear rewriting systems on graph-like structures that are
well-switching.
% and we show that rewriting systems on graphs with
% equivalences are so.
As a sanity check, we also prove that linear rewriting systems on
$\mathcal{M}$-adhesive categories are well-switching.

On the other hand, we argue that a concurrency theory of rewriting can
be developed for well-switching rewriting systems, recovering, sometimes in
weakened form, results which hold in the linear case.
% Specifically, the Church-Rosser theorem is fully recovered.
%
The development has to face a main technical obstacle. In the
literature on linear rewriting systems, a central result is the
characterisation of switch equivalence of sequences of steps based on
what are called processes or consistent
permutations~\cite{CMR:GP,Handbook,BaldanCHKS06,heindel2009category}.
In turn this is the key to derive some
% basic but
fundamental
properties for linear rewriting,
% systems
namely

@@

    A $\mor(\X)$-Van Kampen square will be called simply \emph{Van
      Kampen} and similarly, a $\mor(\X)$-stable square is called
    \emph{stable}.
  \end{definition}

@@

$\mathcal{M}$-adhesivity is well-behaved with respect to  the construction of slice and functor categories \cite{mac2013categories}, as shown by the following theorems from~\cite{ehrig2006fundamentals,lack2005adhesive}.

@@

\begin{example}
  \label{ex:seq-ind}
  Consider a rewriting system in $\cat{Graph}$, the category of
  directed graphs and graph morphisms, which is adhesive. The set of
  rules $\R = \{ \rho_0, \rho_1, \rho_2\}$ consists of the three rules
  of Fig.~\ref{fi:rules}, where numbers are used to represent
  the morphisms from the interface to the left- and right-hand
  sides. Rules $\rho_0$ and $\rho_1$ are linear: $\rho_0$ generates a
  new node, while $\rho_1$ creates an edge. Instead, $\rho_2$ is
  left-linear but not linear, as it ``merges'' two nodes.\todo{P: se $\rho_0$ generasse un arco tra 1 e 2, poi le derivazioni $\der{D}$ e $\der{D}'$ che non vogliamo siano switch equivalent hanno grafi intermedi non isomorfi, non solo con isomorfismi non consistenti, che mi pare meglio. Da vedere.}

@@

We can identify a large class of adhesive categories such that all
left-linear rewriting systems over such categories are strong
enforcing. Such class includes $\cat{Set}$ and it is closed
by comma and functor category constructions.  As such, it includes
essentially all categories (e.g., presheaves over set) that are
typically considered for modelling purposes. Notably, it contains the
category $\cat{Graph}$ of directed graphs.
This is a natural generalisation from adhesive to $\mathcal{M}$-adhesive catgeories of a class studied in \cite{baldan2011adhesivity} (see \Cref{app:fill} for details).

@@

\begin{example}[Non-strong enforcing left-linear rewriting system]
  \label{ex:diff2}
  In light of \Cref{pr:church}, \Cref{ex:diff1} provides an example of
  an independence pair which is not strong. This gives an example of a
  left-linear rewriting system in a $\mathcal{M}$-adhesive category
  which is not strong enforcing. The example is somehow
  pathological since $\mathcal{M}=\mathsf{Iso(\X)}$, but it needs to be so, since natural examples are all in the well-beahved class mentioned above. 
  .\todo{P: Scritto qualcosa, probabilmente da migliorare \\ D: per me va bene così}
  % Instead every
  % left-linear system in category $\cat{Graph}$ is strong enforcing, by
  % the results of Appendix~\ref{app:fill}, summarized
  % in~\Cref{ex:graph-type-bb} and~\Cref{cor:llr-is-strEnf}.
\end{example}

@@

  As a consequence, $\der{D}$ and $\der{D}'$ would be switch
  equivalent, but this is not acceptable when viewing equivalence
  classes of derivations as concurrent computations: in $\der{D}$ the
  first two steps are not sequential independent, while in $\der{D}'$
  they are, intuitively because in $\der{D}$ rule $\rho_1$ uses the
  node generated by $\rho_0$ (adding a self loop to it), while in
  $\der{D}'$ rule $\rho_1$ uses the node that was in the initial
  graph. From the technical point of view, the property of being
  switch equivalent is not closed by prefix, and this prevents
  to derive a sensible concurrent semantics: 
  In fact $\der{D} = \dder{D}_0\cdot \dder{D}_1 \cdot \dder{D}_2$ and
  $\der{D}' = \dder{D}_0'\cdot \dder{D}_1' \cdot \dder{D}_2'$ are switch 
  equivalent, while if we consider the first
  two steps, derivations $\dder{D}_0 \cdot \dder{D}_1$ and
  $\dder{D}_0' \cdot \dder{D}_1'$  are not switch equivalent.
  % if we limit to the first
  %two steps, we obtain derivations $\dder{D}_0 \cdot \dder{D}_1$ and
  %$\dder{D}_0' \cdot \dder{D}_1'$ that are not switch equivalent, while
  %$\der{D} = \dder{D}_0\cdot \dder{D}_1 \cdot \dder{D}_2$ and
  %$\der{D}' = \dder{D}_0'\cdot \dder{D}_1' \cdot \dder{D}_2'$ would
  %be.
\end{example}

@@

A first class consists of those rewriting systems over
possibly hierarchical graphical structures obtained as algebras of
suitable signatures where rules are constrained not to merge elements
of top level sorts in the hierarchy (e.g., for graphs, nodes can be
merged while edges cannot). The idea here is to consider as graphs presheaves on categories in which there are object which plays the roles of \emph{roots}, i.e.~objects which are not codomains of any arrows besides the identity. \todo{D: quest'ultima frase ha senso?}%We first introduce signatures that are
%essentially acyclic graphs whose nodes can be seen as sorts.

\begin{definition}[Root-preserving graphical rewriting systems]
	Let $\X$ be a category, an object $X\in \X$ is a \emph{root} if the only arrow with codomain $X$ is $\id{X}$. The category The
	category $\gph{X}$ of \emph{$\X$-graphs} is the category
	$\Set^{\X}$. A
	\emph{root-preserving graphical rewriting system} is a left-linear
	rewriting system $(\gph{X}, \R)$ such that  the
	following hold for each rule
	$(l\colon K\to L, r\colon K\to R)$ in $\R$
	\begin{enumerate}
		\item for every $X\in \X$ and $x\in L(X)$, there exists a root $Y$
		 and an arrow $f\colon Y\to X$ such that $x$ belongs to the image of
		$L(f)\colon L(Y)\to L(X)$;
		\item $r\colon K\to R$ is mono on the roots, i.e.~for every root $X\in \X$, the component $r_X:K(X)\to R(X)$ is injective.
	\end{enumerate}
\end{definition}

@@

For instance, the category $\cat{Graph}$ can be obtained
by taking as $\X$ the category generated by $E \rightrightarrows V$. In this case $E$ is the only root, hence, condition $1$ asks that in the left-hand side of each rule
there are no isolated nodes, while condition $2$ asks that the
morphism $r: K \to R$ is mono on edges, i.e.~it can only merge nodes.

@@

Left-linear rewriting systems on $\cat{EGraphs}$, i.e.~in which the right half of a rule is mono but not regular in general, 
are well-switching. They have been used in~\cite{BaldanGM06} for the graphical implementation of nominal calculi, where,
differently from~\cite{Gad07}, as a result of name passing the received name is not merged with a local one at the receiver,
but put instead in the same equivalence class, thus better tracing the causal dependencies among processes reductions.
\end{example}

@@

\begin{example}
  \label{ex:non-global}
  Consider a new rewriting system in $\cat{Graph}$ consisting of the
  rules in $\rho_0$, $\rho_1$ and $\rho_2$ in
  Fig.~\ref{fi:rules-cas}.\todo{P: rev.2 dice ``In the top-left and
    top-middle nodes of $\rho_3$, the labels "12" should be "1" and
    "2".''. Non e' corretto, ma magari si capisce poco ...}  Consider
  the derivation $\der{F}$ in Fig.~\ref{fi:derF}. Note that rule
  $\rho_2$ needs the black self-loop on node $12$ to be applied. This can be
  generated either by $\rho_0$ or by $\rho_1$ merging nodes $1$ and
  $2$, hence at least one of them must precede the application of
  $\rho_2$. Indeed, in $\der{F}$ it can be easily seen that the second
  and the third step applying $\rho_1$ and $\rho_2$ are
  independent. However, we could switch twice the application of
  $\rho_0$, bringing it in the last position, obtaining a derivation
  $\der{F}'$ as depicted in Fig.~\ref{fi:derF1}, where the
  applications of $\rho_1$ and $\rho_2$ are no longer independent.

@@

The lemmas above open the way to the key result of this section establishing a canonical form for
switching derivations.

The first result shows that whenever equating two derivation sequences by performing a sequence of switching, we can limit ourselves to switchings inverting the order of steps that in the target derivation has to be reversed, i.e.~we can limit ourselves to applying inversions.

@@

While inversions can be
performed in any order in the case of linear rewriting systems, this is no longer true for left-linear
rewriting systems due to the fact that, as already observed, globality of independence holds in a weak form. For instance, if we get back to Example~\ref{ex:non-global} and consider derivation $\der{F}$ in Fig.~\ref{fi:derF}, the last two steps applying rules $\rho_1$ and $\rho_2$ can be switched, thus leading to a derivation, call it $\der{F}''$,  applying $\rho_0, \rho_2, \rho_1$. Such $\der{F}''$ is switch equivalent to $\der{F}'$ in Fig.~\ref{fi:derF1}, hence starting from $\der{F}''$ we can obtain $\der{F}'$ via a switching sequence, but we cannot start by switching $\rho_0$ and $\rho_2$, despite the fact that this is an inversion.

@@

The class of well-switching $\mathcal{M}$-adhesive rewriting systems is large. Generalising a result in~\cite{baldan2011adhesivity},
we showed that one of its defining properties (the fact that sequential independence implies switchability) holds 
for a general class of $\mathcal{M}$-adhesive categories that include all presheaves over $\cat{Set}$. Moreover, the second property, uniqueness 
of switching, is ensured when rewriting possibly hierarchical graphical structures, as long as elements of the sorts 
of depth 0  are not merged (e.g., one can consider rewriting systems over directed graphs where only nodes are merged, as it 
happens in most of the approaches to the modelling of calculi and biological systems that we cited in the introduction, 
as well as in string diagrams).
%\todo{Giusto, refernences?}

@@

On the practical side, the present paper just surveyed the possibility
of modelling e-graphs. This appears to be quite interesting as
concurrent rewriting on such structures is an active area of
research~\cite{abs-2208-06295}. \todo{F: altre citazioni?} 
We plan to make the correspondence precise
and investigate how the analysis techniques enabled by a concurrent
semantics of rewriting can impact on e-graphs.

@@

\begin{definition}[Consistent permutation]
	\label{def:permcon}
	Let $\X$ be an $\mathcal{M}$-adhesive category and $(\X, \R)$ 
	a left-linear rewriting system.  Take two
	non-empty derivations $\der{D}$ and
	$\der{D}'$ with the same length and with
	isomorphic sources and targets.

	If $\der{D}=\{\dder{D}_i\}_{i=0}^n$ and
	$\der{D}'=\{\dder{D}'_i\}_{i=0}^n$ with associated sequence of rules
	$r(\der{D})=\{\rho_i\}_{i=0}^n$ and
	$r(\der{D}')=\{\rho'_i\}_{i=0}^n$, a \emph{consistent permutation}
	between $\der{D}$ and
	$\der{D}'$ is a permutation
	$\sigma\colon [0,n]\to [0,n]$ such that, for every $i\in [0,n]$,
	$\rho_i=\rho'_{\sigma(i)}$ and, moreover, there exists a
	\emph{mediating isomorphism}
	$\xi_\sigma\colon \tpro{D} \to \lpro \der{D}' \rpro$ fitting in the
	following diagrams, where $m_i, m'_i, h_i$ and $h'_i$ are,
	respectively, the matches and co-matches of $\dder{D}_i$ and
	$\dder{D}'_i$.
	\[
	\xymatrix@C=30pt{L_i
		\ar[r]^{m_i} \ar[d]_{m'_{\sigma(i)}}& G_i \ar[r]^{\iota_{G_i}}
		&\tpro{D} \ar[d]^{\xi_\sigma} & R_i \ar[r]^{h_i}
		\ar[d]_{h'_{\sigma(i)}}& G_{i+1} \ar[r]^{\iota_{G_{i+1}}}
		&\tpro{D} \ar[d]^{\xi_\sigma} \\G'_{\sigma(i)}
		\ar[rr]_{\iota'_{G'_{\sigma(i)}}}&& \lpro \der{D}' \rpro&
		G'_{\sigma(i)+1} \ar[rr]_{\iota'_{G'_{\sigma(i)+1}}}&& \lpro
		\der{D}' \rpro}
	\]
\end{definition}
We can now turn to the proofs of the results of \Cref{subsec:canonical}.
